---
title: "Activity Monitoring Analysis"
author: "Daniel Smit"
date: '2022-06-19'
output:
  html_document:
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

We load the data from a CSV.

```{r loadingCSV}
activity = read.csv('activity.csv');

summary(activity)
```

Notice a number of NA step values.
Until we start estimating the NA values, we will exclude results with NA values.

```{r removingNA}
activityNoNA = na.omit(activity)
```

## What is mean total number of steps taken per day?

A histogram of all values shows that zero steps were recorded on a large proportion of days.

```{r stepsHistogram}
stepsPerDay = aggregate(steps ~ date, activityNoNA, sum)

hist(x=stepsPerDay$steps, breaks=30,
     main='Histogram of Steps Taken Per Day',
     xlab='Steps Taken Per Day',
     ylab='Count')
```

We calculate the mean and median as follows.

```{r calculateMeanAndMedian}
meanSteps = mean(stepsPerDay$steps)
medianSteps = median(stepsPerDay$steps)

cat('Mean steps: ', meanSteps, '\n')
cat('Median steps: ', medianSteps)
```

## What is the average daily activity pattern?

First we determine the mean steps for each 5 minute interval, then display on a time series plot.

```{r calcuklateMeanByInterval}
meanStepsByInterval = aggregate(steps ~ interval, activityNoNA, mean)

library(ggplot2)
ggplot(meanStepsByInterval, aes(x=interval, y=steps)) + geom_line() +
  ggtitle("Mean Steps Per 5 Minute Interval")

```

What interval contains the max number of steps on average?

```{r calculateMaxMeanStepInterval}
rowWithMaxMeanSteps = meanStepsByInterval[which.max(meanStepsByInterval$steps),]

cat('Max mean steps during interval ', rowWithMaxMeanSteps$interval, ' with ', rowWithMaxMeanSteps$steps, ' steps.')
```

## Inputting missing values

How many rows contain NAs?

```{r numberOfRowsContainingNAs}
numRowsWithNAs = sum(!complete.cases(activity))

cat(numRowsWithNAs, ' rows with NAs')
```

Let's bring the number of NA step values to zero through estimation, by using the mean steps for that time interval across the entire series.

For this, let's continue using the `meanStepsByInterval` as calculated earlier.

```{r fillNAs}
intervalValues <- unique(activity$interval)

filledActivity = activity
stepsCol <- 1
intervalCol <- 3

for(row in 1:nrow(filledActivity)){
  if(is.na(filledActivity[row,stepsCol])) {
    intervalIndex <- match(filledActivity[row,intervalCol], intervalValues)
    filledActivity[row,stepsCol] <- meanStepsByInterval[intervalIndex, 2]
  }
}

summary(filledActivity)
```

Let's plot a histogram of this estimated data.

```{r stepsHistogramWithEstimatedData}
stepsPerDayEstimated = aggregate(steps ~ date, filledActivity, sum)

hist(x=stepsPerDayEstimated$steps, breaks=30,
     main='Histogram of Steps Taken Per Day with estimation',
     xlab='Steps Taken Per Day',
     ylab='Count (with estimation)')
```

Let's calculate the mean and median.

```{r calculateFilledMeanAndMedian}
filledMeanSteps = mean(stepsPerDayEstimated$steps)
filledMedianSteps = median(stepsPerDayEstimated$steps)

cat('Filled mean steps: ', filledMeanSteps, '\n')
cat('Filled median steps: ', filledMedianSteps)
```

Just to remind you once again, the original mean and medians, before filling in NA step values, were:

```{r printOriginalMeanAndMedianAgain}
cat('Mean steps: ', meanSteps, '\n')
cat('Median steps: ', medianSteps)
```

The means are the same before and after filling in missing data, but the median has changed slightly.

There are also slight differences in the histograms, with the count of the most common step range being amplified and surrounding values lowered.

## Are there differences in activity patterns between weekdays and week-ends?

Let's go back to the original data set and add a column specifying whether the row was taken on a weekend or not.

```{r printOriginalMeanAndMedian}
library(chron)
library(dplyr)

activityWithWeekendCol = mutate(activity, isWeekend = is.weekend(date))

```

Now let's plot the time series plots for weekdays and weekends.

```{r displayWeekendAndWeekdayPlots}
weekdayActivity = activityWithWeekendCol[!activityWithWeekendCol$isWeekend,]
weekendActivity = activityWithWeekendCol[activityWithWeekendCol$isWeekend,]

meanStepsByIntervalOnWeekdays = aggregate(steps ~ interval, weekdayActivity, mean)
meanStepsByIntervalOnWeekends = aggregate(steps ~ interval, weekendActivity, mean)

weekdaysPlot <- ggplot(meanStepsByIntervalOnWeekdays, aes(x=interval, y=steps)) + geom_line() +
  ggtitle("Mean Steps Per 5 Minute Interval on Weekdays")

weekendsPlot <- ggplot(meanStepsByIntervalOnWeekends, aes(x=interval, y=steps)) + geom_line() +
  ggtitle("Mean Steps Per 5 Minute Interval on Weekends")

require(gridExtra)
grid.arrange(weekendsPlot, weekdaysPlot, nrow=2)

```

We notice a fairly significant difference between the weekends and weekday means. On weekdays, there was less activity from interval 1000 onwards than on weekends. Both plots do however show a lack of activity during late night/early morning hours, and a stretch of increased activity just prior to interval 1000.